---
title: "Tephritid species identification using CNN"
author: "Aaron Harmer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message = FALSE, warning = FALSE}
# library(reticulate)
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, patchwork)

dat = read_csv("C:/Users/HarmerA/OneDrive - MWLR/Documents/data/tephritidML/models/tephritid_species_V3_3_Xception_200_epoch_transfer_earlyStop_log.csv") %>% 
  pivot_longer(!epoch, names_to = "stat", values_to = "value")
```

### Summary
A convolution neural network was trained via transfer learning to identify Tephritidae species using wing images. Threefold cross-validation was used to achieve an accuracy of `r round(dat[nrow(dat)-1,3] * 100)`% at predicting class. The initial image set was small, comprising 674 images (459 training, 215 validation) in  30 classes. Classes were defined by genus name or genus & species name.

<br>

### Image pre-processing
Images were prepared by:

* Converting colour images to grayscale
* Blurring to reduce image grain noise
* Standardising brightness and contrast to mean = 0.5, contrast range = +-2 standard deviations, with the extreme values clipped
* Squashing and downsampling to 299 by 299 pixels to match the network input filter size

<br>

### Model training
Transfer learning was used to retrain an Xception model initially trained on the Imagenet dataset. The model was retrained initially for 148 epochs (early-stopping set) at a learning rate of 1e-04, and then fine-tuned for a further 77 epochs (early-stopping set) at a learning rate of 1e-05.

Images were augmented to prevent over-fitting and increase the robustness of the model to variation in additional images added later. Images augmentations included:

* Random horizontal and vertical shift up to 10%
* Random zoom up to +- 10%
* Random rotation up to +- 25 degrees

<br>

### Results

The model achieved a validation accuracy of `r round(dat[nrow(dat)-1,3] * 100)`% and validation loss of `r round(dat[nrow(dat),3], digits = 2)` after 77 epochs (Figure 1).

<br>

```{r warning = FALSE, fig.width = 12, fig.cap = "Fig. 1: Accuracy and loss of Xception model after 77 epochs.", fig.align="center"}
acc_plot = dat %>% 
  filter(stat == "accuracy" | stat == "val_accuracy") %>% 
  ggplot(aes(x = epoch, y = value, colour = stat)) +
    geom_line(linewidth = 1) +
    scale_colour_manual(labels = c("Training accuracy", "Validation accuracy"), values = c("#70ad47", "#4472c4")) +
    ylim(c(0,1)) +
    xlab("\nEpoch") +
    ylab("Accuracy\n") +
    theme_classic(base_size = 16) +
    theme(legend.title = element_blank(), legend.position = c(0.75, 0.25))

loss_plot = dat %>% 
  filter(stat == "loss" | stat == "val_loss") %>% 
  ggplot(aes(x = epoch, y = value, colour = stat)) +
    geom_line(linewidth = 1) +
    scale_colour_manual(labels = c("Training loss", "Validation loss"), values = c("#70ad47", "#4472c4")) +
    ylim(c(0,1)) +
    xlab("\nEpoch") +
    ylab("Loss\n") +
    theme_classic(base_size = 16) +
    theme(legend.title = element_blank(), legend.position = c(0.75, 0.75))

acc_plot|loss_plot
```




``` {r}
idx = which.max(py$preds)
labels = as.vector(unlist(read_csv("C:/Users/HarmerA/OneDrive - MWLR/Documents/data/tephritidML/tephritid_annotation.csv", col_names = FALSE)))
name = labels[idx]
p = paste0(round(max(py$preds) * 100, digits = 1), "%")

plot(rast(wing), col = gray.colors(256), asp = 1, legend = FALSE, axes = FALSE)
plot(rast(hmap), legend = FALSE, axes = FALSE, col = viridis(256), alpha = 0.4, add = TRUE)
text(x = 1, y = 285, labels = name, pos = 4, cex = 1)
text(x = 1, y = 270, labels = p, pos = 4, cex = 1)


table of misIDs
```
